CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to Support Active Travel}

### 2023-10-03 paper.qmd
Figure caption

![](images/paste-5.png)

![](images/paste-6.png)

>>>>

<!-- Add figure with a single row and subcaptions for Fastest and Quietest route types: -->

::: {#fig-route-types}

!['Fastest' route network for Limerick](images/paste-5.png)

!['Quietest' route network for Limerick](images/paste-6.png)

Illustration of the 'Fastest' and 'Quietest' route networks for central and East Limmerick.

:::

### 2023-10-03 paper.tex
Figure caption

\begin{frontmatter}
\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source
Methodology to Support Active Travel} }
\author[1]{Robin Lovelace%
\corref{cor1}%
}
 \ead{r.lovelace@leeds.ac.uk} 
\author[1]{Joey Talbot%
%
}

@@

To address data needs for both safety, business case development and
project evaluation on the National Road Network (NRN), the Cycle Route
Uptake and Scenario Estimation (CRUSE) Tool was developed. Cycling in
Ireland represents only 3\% of total modal share as of the latest Census
(2016), but accounts for 20\% of serious injuries and 7\% of all
fatalities. High resolution datasets on baseline and potential cycling
levels are needed to meet Ireland's climate and safety targets.

Due to an absence of reliable data, particularly rural cycling levels,
TII commissioned the Institute for Transport Studies (ITS) at the
University of Leeds and AECOM to develop a new tool for this purpose.
ITS Leeds led the development of the PCT for England and Wales, which
has ``revolutionized the practice of strategic cycle planning
nationally''. The tool is an open-source approach using recognized
open-source methodology to enable planners, engineers, and other
stakeholders to make evidence-based decisions for the NRN. CRUSE is
available at \url{https://cruse.bike/} and builds on the Propensity to
Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
several important ways, modeling denser networks at higher levels of
geographic resolution and additional purposes, including simulated
recreational, social, shopping and personal trips, in addition trips to
work and primary, secondary and tertiary education locations. The tool
provides key `denominator' information to enable reporting of collision
rates per billion km cycled.

CRUSE is structured in a similar way to the traditional four-stage
transport model, but its use of Open Street Map (OSM) data, used by
\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
enables network quality to be assessed without costly surveys to record
new infrastructure. OSM tags generate ``cycle friendliness'' estimates
of all links on the network, based on existing recorded infrastructure.
Many networks are provided, highlighting routes for directness (Fastest)
and ``cycle friendliness'' (Quietest). It uses origin and destination
data from the 2016 Census in combination with modeled demand data to
estimate current and potential future cycling levels at area, route, and
network levels for each county in Ireland.

For investment in cycling to deliver on the huge potential benefits,
cycling must become safer. With growth in the E-bike market, the tool
will help inform inter-urban and rural networks to support the transfer
of trips to sustainable modes for longer journeys. The findings and
resulting tool are directly relevant to addressing the challenges and
opportunities faced by National Road Authorities (NRAs) across Ireland.
The provision of open access evidence, that can be used non-experts and
professionals, is relevant to countries with sustainable transport
policies worldwide.
\end{abstract}

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

Mobility kills. High energy transport systems are a major contributor to
climate change, a leading cause of premature death and injury due to
road traffic collisions, and a cause of disease due to airborne
pollutants. Conversely, evidence-based and effective transport
\emph{policies} have great potential to decarbonize economies, improve
public health, and save lives.

Transport is responsible for 23\% of global greenhouse gas emissions,
70\% of which are from road transport, nearly half of which (around 10\%
of global emissions) is from passenger cars \citep{jaramillo2022}. The
transport system encourages, enables and in some cases enforces
unsustainable lifestyles. Services that are only accessible by car
lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The framework highlights the importance of demand
reduction (\emph{avoid}ing unnecessary trips), in addition to mode
\emph{shift} to sustainable modes and \emph{improvement} of existing
energy converters, in that order.

@@

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
and the National Roads 2040 strategy\footnote{
  https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
These policy frameworks are backed-up by public support for active
travel interventions: while only 3\% of trips in Ireland are made by
cycling according to the latest Census data from 2016,\footnote{
  https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
``83\% of respondents supported enhanced walking and cycling
facilities''.

At the regional level, the recently published Greater Dublin Area
Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
reinforces these findings: ``nearly a quarter of adults cycle at least
once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
area taking up to 60,000 cars off the road today. Extrapolating this on
a per population basis across Ireland, with around 40\% of the
population living in Dublin, this suggests that around 150,000 cars
could be removed nationwide just by achieving Dublin levels of cycling
in all counties (notwithstanding existing cycling trips and differences
in trip distances). 7 in 10 trips in Ireland are by car\footnote{
  https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}.
Cycling has the potential to replace a large proportion of these trips:
``A high priority must also be given to cyclists, because trips by this
mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as e-bikes extend the range of this mode''\footnote{
  https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.

@@

A key consideration for TII is to ensure both urban and rural Ireland
were targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process. Other use cases in the Irish context
includes monitoring the need for existing cycling infrastructure
upgrades, undertaking appraisal for business case development, as well
as safety and exposure information. The latter is in line with the
European Cycling Federation, Dutch Cycling Embassy recommendations and
the European RISM Directive, for reporting collision rates of vulnerable
road users by 2024. These factors mean that the tool can be seen as an
open access `leverage point' in the planning system
\citep{lovelace2020}.

@@

The rest of paper is structured as follows. In
Section~\ref{sec-methods}, we outline the methods used to generate the
evidence presented in the CRUSE Tool for Ireland. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

@@

The first version of the PCT was based on current and future potential
uptake of cycling for single stage \emph{travel to work} at desire line,
zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning, as part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modelling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

>>>>

\begin{frontmatter}
\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to
Support Active Travel} }
\author[1]{Robin Lovelace%
\corref{cor1}%
}
 \ead{r.lovelace@leeds.ac.uk} 
\author[1]{Joey Talbot%
%
}

@@

The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was
commissioned by TII and developed by the Leeds Institute for Transport
Studies (ITS) and AECOM To address data needs for safety, business case
development and project evaluation on the National Road Network (NRN).
Cycling in Ireland represents only 3\% of total modal share as of the
latest Census (2016), but accounts for 20\% of serious injuries and 7\%
of all fatalities. High resolution datasets on baseline and potential
cycling levels are needed to meet Ireland's climate and safety targets.

The CRUSE project builds on the methods underlying the Propensity to
Cycle Tool for England and Wales (PCT), which had a major impact on
strategic cycle network planning in the UK. Like the PCT, CRUSE is based
on open source software and reproducible data science. The resulting web
application is available at \url{https://cruse.bike/}, enabling
planners, engineers, and other stakeholders to make more evidence-based
decisions for the NRN. CRUSE goes beyond the PCT in several important
ways, modeling denser networks at higher levels of geographic resolution
and additional purposes, including simulated recreational, social,
shopping and personal trips, in addition to trips to work and primary,
secondary and tertiary education locations. The tool provides key
`denominator' information to enable reporting of collision rates per
billion km cycled.

Another key feature of CRUSE is its use of Open Street Map (OSM) data,
which is continuously updated, and used by
\href{https://www.cyclestreets.net/}{Cyclestreets.net} for route
calculation and ``cycle friendliness'' estimates of all links on the
network. CRUSE provides three network types: the `Fastest' network
highlights routes for directness; `Quietest' prioritises cycle
friendliness, and `Balanced' represents a balance between speed and
comfort. Trips are generated based on origin and destination data from
the 2016 Census in combination with modeled demand data to estimate
current and potential future cycling levels at area, route, and network
levels for each county in Ireland.

For investment in cycling to deliver on the huge potential benefits,
cycling must become safer. With growth in the E-bike market, CRUSE will
help inform inter-urban and rural networks to support the transfer of
trips to sustainable modes for longer journeys. Provision of open access
data in a publicly available web tool will address the challenges faced
by National Road Authorities (NRAs) across Ireland in meeting their
sustainability commitments. The experience of developing a deploying
CRUSE are relevant to countries with sustainable transport policies
worldwide.
\end{abstract}

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, boxrule=0pt, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

High energy transport systems are a major contributor to climate change,
a leading cause of premature death and injury due to road traffic
collisions, and a cause of disease due to airborne and noise pollutants.
Conversely, evidence-based and effective transport \emph{policies} have
great potential to decarbonize economies, improve public health, and
save lives.

Transport is responsible for 23\% of global greenhouse gas emissions,
70\% of which are from road transport. Nearly nearly half of these
emissions (around 10\% of global emissions) are from passenger cars
\citep{jaramillo2022}. The transport system encourages, enables and in
some cases enforces unsustainable lifestyles. Services that are only
accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The ASI framework highlights the importance of
demand reduction (\emph{avoid}ing unnecessary trips), in addition to
mode \emph{shift} to sustainable modes and \emph{improvement} of
existing energy converters, in that order.

@@

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
and the National Roads 2040 strategy \footnote{https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
These policy frameworks are backed-up by public support for active
travel interventions: while only 3\% of trips in Ireland are made by
cycling according to the latest Census data from 2016,\footnote{https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
``83\% of respondents supported enhanced walking and cycling facilitie
s''.

At the regional level, the recently published Greater Dublin Area
Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
reinforces these findings: ``nearly a quarter of adults cycle at least
once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
area taking up to 60,000 cars off the road today. Extrapolating this on
a per population basis across Ireland, with around 40\% of the
population living in Dublin, this suggests that around 150,000 cars
could be removed nationwide just by achieving Dublin levels of cycling
in all counties (notwithstanding existing cycling trips and differences
in trip distances). 7 in 10 trips in Ireland are by car\footnote{https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}
. Cycling has the potential to replace a large proportion of these
trips: ``A high priority must also be given to cyclists, because trips
by this mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as e-bikes extend the range of this mode'' \footnote{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.

@@

A key consideration for TII is to ensure both urban and rural Ireland
were targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process.

TII also requires evidence to support monitoring of growth in cycling,
upgrades of existing cycling infrastructure, business case development
and appraisal, and exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.

@@

The main aim of the paper is to present the CRUSE Tool and describe the
methods and datasets on which the estimates of current and future
potential cycling levels are based. We will also highlight key results,
discuss the implications for policy and practice, and outline how
similar tools could be developed in other countries. It is structured as
follows. In Section~\ref{sec-methods}, we outline the methods used to
generate the evidence presented in the CRUSE Tool for Ireland. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

@@

The first version of the PCT was based on current and future potential
uptake of cycling for single stage \emph{travel to work} at desire line,
zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning and part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modelling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

\hypertarget{routing}{%
\subsection{Routing}\label{routing}}

Routes in CRUSE are generated by CycleStreets, a not-for profit
transport consultancy and web development company who provide
application programming interfaces (APIs) supplying a range of datasets
for cycle planning and advocacy. Although based in Cambridge, UK,
CycleStreets provide routing services `by cyclists, for cyclists'
internationally, including in Ireland. The CycleStreets routing engine
is based on OpenStreetMap (OSM) data, which is continuously updated by a
global community of volunteers.

CycleStreets' services can be used by anyone for free, but we
commissioned a bespoke routing service for CRUSE for the following
reasons, to enable:

\begin{itemize}
\tightlist
\item
  Calculation of hundreds-of-thousands of routes, which is against the
  terms of service of the free API.
\item
  Making changes to the routing profiles, including allowing routing on
  trunk roads, which are not permitted in the default routing profiles.
\item
  Control over the version of OSM data being used for the routing,
  allowing regular updates to the route networks as OSM data is updated.
\end{itemize}

We calculated three network types for each disaggregated (`jittered') OD
pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
CycleStreets'
\href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
the `Fastest' routes minimise journey time, accounting for traffic
lights and surface type. The `Quietest' route type minimises busy
sections of road, allowing high `diversion factors' from the fastest
route on quieter but less direct ways (often avoiding roads and
interactions with motor vehicles altogether where possible). The
`Balanced' route type is a compromise between the two, minimising
journey time while avoiding the busiest roads. Allowing users to switch
between these three route types enables them to consider the trade-offs
between directness and `cycle friendliness' when planning new
infrastructure, as shown in Figure~\ref{fig-route-types}.

{\centering \includegraphics{images/paste-5.png}

\caption{`Fastest' route network for Limerick}

{\centering \includegraphics{images/paste-6.png}

\caption{`Quietest' route network for Limerick}

\caption{\label{fig-route-types}Illustration of the `Fastest' and
`Quietest' route networks for central and East Limmerick.}

\hypertarget{sec-scenarios}{%
\subsection{Scenarios of cycling potential}\label{sec-scenarios}}

The datasets outlined above were used to generate estimates of cycling
currently (the `Baseline' scenario) and under four scenarios of cycling
potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
of these scenarios is described below.

The Baseline scenario approximates current cycling levels. As outlined

Near market Climate Action Plan Go Dutch Ebike
